---
title: "High-SDI alignment counterfactual"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
source(here::here("scripts/regional_analysis/R/packages.R"))
source(here::here("scripts/regional_analysis/R/data_loaders.R"))
source(here::here("scripts/regional_analysis/R/metrics.R"))
source(here::here("scripts/regional_analysis/R/plots.R"))
source(here::here("scripts/simulation-sdi-response/high_sdi_counterfactual.R"))

simulation_seed <- 20260615
n_candidates <- 50000
n_keep <- 5000
median_abs_error_warn <- 0.02
max_abs_error_warn <- 0.05

# Faster exploratory setting:
# n_candidates <- 5000
# n_keep <- 500
```

## Summary

This report tests whether lower-SDI attention-burden alignment can be explained
as a passive consequence of the process that produces observed High SDI
alignment plus shared disease burden. The null expectation is that lower-SDI
concentration-index trends should track the High-SDI-calibrated simulation
ribbon. Under that interpretation, any apparent lower-SDI alignment would be
dictated by the determinants of GWAS prioritisation in High SDI settings and by
overlap in disease-burden rankings across SDI regions.

The main finding is that lower-SDI regions generally perform worse than expected
under this transfer null in the time-varying attention analysis, for both the
active and reversed concentration-index definitions. This suggests that observed
time-varying GWAS attention is not representative of the broader set of
allocations compatible with High SDI alignment. It is concentrated on a subset
of High-SDI-compatible conditions that happen to be less compatible with
lower-SDI burden than alternative High-SDI-calibrated allocations would be.

The all-time analysis gives a more nuanced result. When asking whether lower-SDI
disease burden is concentrated among conditions that have received GWAS
attention, all-time attention is closer to the null expectation. When asking
whether GWAS attention is concentrated on lower-SDI high-burden conditions,
lower-SDI regions again perform worse than expected. This suggests some
accumulated coverage of lower-SDI-relevant conditions, but insufficient
prioritisation intensity toward the conditions that matter most in lower-SDI
settings.

## Aim

This analysis tests whether lower-SDI attention-burden alignment is compatible
with a High-SDI transfer null. Under this null, the process that produces
observed High SDI alignment is treated as sufficient to explain lower-SDI
alignment once shared burden structure is accounted for. This process may
include High SDI disease burden, tractability, statistical power, cohort
availability, historical research momentum, and international collaborations;
the simulation does not distinguish those mechanisms.

The simulation is a calibrated permutation analysis. It preserves each GWAS
attention window's observed sparsity and attention-score inequality, breaks the
observed mapping between conditions and attention scores, then keeps simulated
attention allocations that reproduce the observed High SDI alignment metric.
Those same kept allocations are evaluated against each SDI region's burden
ranking.

## Methodology

### Analysis universe

The main analysis uses sliding-window GWAS attention scores and SDI burden data
for all ages, both sexes, and DALY counts. The condition universe is the same
condition set used by the regional concentration-index analysis after zero
padding attention windows. Conditions with zero GWAS attention remain in the
analysis universe.

COVID-19 is excluded before constructing observed metrics or simulated
allocations. The exclusion is used because a single pandemic-era condition can
dominate the support of possible permutations, narrowing the counterfactual
range in a way that is not useful for testing the broader GWAS condition canon.
Because COVID-19 is not uniformly high-burden across SDI regions, this exclusion
may be conservative for lower-SDI below-null findings in some years.

### Metrics

The primary metric is the attention-burden concentration index:

```text
CI_active = CI(GWAS attention ranked by DALY burden)
```

Positive values mean GWAS attention is concentrated among higher-burden
conditions in the selected SDI region and year.

The reversed metric is calculated separately as a diagnostic:

```text
CI_reversed = CI(DALY burden ranked by GWAS attention)
```

Positive values mean disease burden is concentrated among conditions with
greater GWAS attention. Because this answers a different question, the active
and reversed metrics are calibrated in separate simulation runs.

### Candidate generation

For each year or attention window, the simulation starts with the observed
vector of GWAS attention scores:

```text
observed_scores = {0, 0, ..., nonzero attention scores}
```

Each candidate allocation preserves the exact observed nonzero score values and
the exact number of zero-attention conditions. A candidate is generated by:

```text
1. sample n_nonzero conditions from the condition universe
2. assign the observed nonzero scores to those selected conditions
3. assign zero attention to all other conditions
```

The candidate generator uses a proposal mixture so that enough
High-SDI-aligned allocations are available for calibration:

```text
proposal 1:
  randomly choose nonzero conditions
  randomly assign nonzero score magnitudes

proposal 2:
  randomly choose nonzero conditions
  assign larger scores to higher High SDI burden ranks, with rank noise

proposal 3:
  preferentially choose higher High SDI burden conditions
  assign larger scores to higher High SDI burden ranks, with rank noise
```

The rank-noise levels range from strongly High-SDI-shaped to weakly
High-SDI-shaped. The weighted-selection proposals make higher High SDI burden
conditions more likely to receive nonzero attention before score magnitudes are
assigned. The proposal mixture is not itself the estimand; it is a practical
sampler used to discover enough candidate allocations near the observed High
SDI calibration target.

### Calibration

For each year, the simulation generates `r n_candidates` candidate allocations
and keeps the `r n_keep` allocations closest to the observed High SDI metric.
The active and reversed analyses use different calibration targets:

```text
active run:
  keep allocations closest to observed High SDI CI_active

reversed run:
  keep allocations closest to observed High SDI CI_reversed
```

High SDI is therefore a calibration check, not an independent test. The lower
SDI regions are the test regions: the High-SDI-calibrated allocations are
projected onto each lower-SDI burden ranking to estimate the null distribution
expected from shared burden structure alone.

### Summaries and interpretation

For each SDI region and year, the kept allocations define a null distribution.
The report summarizes:

```text
sim_lci      = 2.5th percentile of simulated CI
sim_median   = median simulated CI
sim_uci      = 97.5th percentile of simulated CI
observed_ci  = empirical regional CI
```

If the observed lower-SDI CI lies inside the null ribbon, the observed regional
alignment is compatible with the High-SDI transfer null. If it lies below the
ribbon, observed alignment is worse than expected under a generic
High-SDI-calibrated canon. If it lies above the ribbon, observed alignment is
stronger than expected under the transfer null.

This is a sufficiency test, not a causal proof. Compatibility with the null
means that the observed pattern can be reproduced by High SDI alignment plus
shared burden structure. It does not prove that High SDI prioritisation caused
the observed pattern.

### Interpretation guide

The lower-SDI observed line is interpreted relative to the High-SDI-calibrated
null ribbon:

```text
observed inside null ribbon:
  lower-SDI alignment is compatible with passive High-SDI transfer

observed above null ribbon:
  lower-SDI alignment is stronger than expected
  possible evidence of condition prioritisation beyond passive High-SDI transfer

observed below null ribbon:
  lower-SDI alignment is weaker than expected
  High-SDI-shaped prioritisation is worse than a generic High-SDI-calibrated
  allocation for lower-SDI burden
```

The last case is especially informative. It means the empirical GWAS canon is
not representative of the broader High-SDI-compatible allocation space. Among
allocations that could satisfy the High SDI calibration target, observed GWAS
attention is concentrated on a subset of conditions that are relatively less
important in lower-SDI regions than other plausible High-SDI-aligned
alternatives.

### Calibration diagnostics

Calibration quality is reported for every year. Windows are flagged when:

```text
median_abs_error > 0.02
or
max_abs_error > 0.05
```

A warning indicates that the kept allocations do not closely reproduce the
High SDI calibration target and should not be used for strong inference without
increasing the candidate count or tuning the proposal sampler. Warning windows
remain visible in plots using a different point shape.

### Sliding-window and all-time analyses

The active and reversed metrics are calibrated in separate simulation runs. The
active ribbon keeps allocations closest to observed High SDI attention-burden
concentration index; the reversed ribbon keeps allocations closest to observed
High SDI reversed concentration index.

The all-time sensitivity uses the all-time attention score distribution and
evaluates it against each burden year.

## Data

```{r prepare-data}
window_data <- prepare_counterfactual_windows(exclude_covid = TRUE)
observed_alignment <- compute_observed_counterfactual_alignment(window_data)

years <- sort(unique(window_data$year))

tibble(
  first_year = min(years),
  last_year = max(years),
  n_years = length(years),
  n_conditions = n_distinct(window_data$cause_id),
  n_locations = n_distinct(window_data$location_name)
)
```

```{r observed-summary}
observed_alignment %>%
  filter(
    sex_name == "Both",
    year %in% c(min(years), 2010, 2015, 2020, max(years)),
    location_name %in% c("High SDI", "Low SDI", "Global")
  ) %>%
  select(
    location_name,
    year,
    concentration_index,
    concentration_index_rev,
    attention_gini,
    alignment_ratio,
    share_alignment,
    n_causes
  ) %>%
  arrange(location_name, year)
```

## Simulation

```{r run-counterfactual}
active_counterfactual <- run_high_sdi_counterfactual(
  window_data = window_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "active",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

active_counterfactual_summary <- summarise_counterfactual(
  simulations = active_counterfactual$simulations,
  observed_alignment = observed_alignment,
  calibration = active_counterfactual$calibration
)

reversed_counterfactual <- run_high_sdi_counterfactual(
  window_data = window_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "reversed",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

reversed_counterfactual_summary <- summarise_counterfactual(
  simulations = reversed_counterfactual$simulations,
  observed_alignment = observed_alignment,
  calibration = reversed_counterfactual$calibration
)
```

```{r calibration-summary}
bind_rows(
  active_counterfactual$calibration,
  reversed_counterfactual$calibration
) %>%
  select(
    year,
    calibration_metric,
    n_candidates,
    n_kept,
    n_nonzero_attention,
    n_zero_attention,
    n_unique_nonzero_condition_sets,
    observed_high_sdi_ci,
    observed_high_sdi_reversed_ci,
    mean_kept_high_sdi_ci,
    mean_kept_high_sdi_reversed_ci,
    observed_calibration_ci,
    mean_kept_calibration_ci,
    median_abs_error,
    max_abs_error,
    calibration_quality,
    seed
  ) %>%
  arrange(calibration_metric, year)
```

Windows are flagged when `median_abs_error` exceeds
`r median_abs_error_warn` or `max_abs_error` exceeds
`r max_abs_error_warn`. A warning marks a window as poorly calibrated for strong
interpretation; it does not remove the window from the plot.

## Observed vs null ribbon

The red line and points show the observed metric. The blue line shows the
median of the High-SDI-calibrated null distribution; the blue ribbon shows its
2.5th to 97.5th percentile range. `observed_percentile_rank` is the proportion
of simulated values less than or equal to the observed value; values near 0
indicate unusually weak observed alignment relative to the null.

```{r null-ribbon-active, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(
  active_counterfactual_summary,
  metric = "active",
  title = "Sliding-window attention: active CI"
)
```

```{r null-count-table-active}
summarise_null_compatibility(active_counterfactual_summary, metric = "active") %>%
  arrange(location_name)
```

```{r null-summary-table-active}
active_counterfactual_summary %>%
  filter(
    year %in% c(min(years), 2010, 2015, 2020, max(years)),
    location_name != "Global"
  ) %>%
  select(
    location_name,
    year,
    observed_ci,
    sim_lci,
    sim_median,
    sim_uci,
    observed_minus_sim_median,
    observed_percentile_rank,
    null_compatibility
  ) %>%
  arrange(location_name, year)
```

## Reversed observed vs null ribbon

The reversed metric asks whether disease burden is concentrated among
conditions with greater GWAS attention. It is included as a diagnostic rather
than the primary attention-burden alignment metric.

```{r null-ribbon-reversed, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(
  reversed_counterfactual_summary,
  metric = "reversed",
  title = "Sliding-window attention: reversed CI"
)
```

```{r null-count-table-reversed}
summarise_null_compatibility(reversed_counterfactual_summary, metric = "reversed") %>%
  arrange(location_name)
```

```{r null-summary-table-reversed}
reversed_counterfactual_summary %>%
  filter(
    year %in% c(min(years), 2010, 2015, 2020, max(years)),
    location_name != "Global"
  ) %>%
  select(
    location_name,
    year,
    observed_reversed_ci,
    sim_reversed_lci,
    sim_reversed_median,
    sim_reversed_uci,
    observed_reversed_minus_sim_median,
    observed_reversed_percentile_rank,
    reversed_null_compatibility
  ) %>%
  arrange(location_name, year)
```

## Calibration check

High SDI is the calibration target. Its observed line should sit close to the
kept permutation distribution by construction; lower-SDI regions are the actual
test of the transfer null. Warning symbols mark years where the retained
permutations did not closely reproduce the High SDI calibration target.

```{r high-sdi-active-calibration-plot, fig.width=7, fig.height=4}
plot_counterfactual_ribbon(
  active_counterfactual_summary %>% filter(location_name == "High SDI"),
  include_global = TRUE,
  metric = "active",
  title = "High SDI calibration check: active CI"
)
```

```{r high-sdi-reversed-calibration-plot, fig.width=7, fig.height=4}
plot_counterfactual_ribbon(
  reversed_counterfactual_summary %>% filter(location_name == "High SDI"),
  include_global = TRUE,
  metric = "reversed",
  title = "High SDI calibration check: reversed CI"
)
```

## Effective sparsity

```{r sparsity-plot, fig.width=7, fig.height=4}
active_counterfactual$calibration %>%
  ggplot(aes(x = year, y = n_nonzero_attention)) +
  geom_line() +
  geom_point() +
  theme_report() +
  labs(
    x = "GWAS attention window",
    y = "Nonzero attention conditions"
  )
```

## All-time attention sensitivity

This sensitivity repeats the counterfactual using all-time GWAS attention scores
and each year of DALYs. It asks whether the accumulated GWAS condition canon is
narrower than a generic High-SDI-calibrated canon across the burden history.
Because the all-time attention distribution is dense and highly unequal,
calibration warnings should be checked before placing strong weight on these
ribbons under the default candidate count.

```{r all-time-counterfactual}
all_time_data <- prepare_counterfactual_all_time(
  burden_years = years,
  exclude_covid = TRUE
)

all_time_observed_alignment <- compute_observed_counterfactual_alignment(all_time_data)

all_time_active_counterfactual <- run_high_sdi_counterfactual(
  window_data = all_time_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "active",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

all_time_active_summary <- summarise_counterfactual(
  simulations = all_time_active_counterfactual$simulations,
  observed_alignment = all_time_observed_alignment,
  calibration = all_time_active_counterfactual$calibration
)

all_time_reversed_counterfactual <- run_high_sdi_counterfactual(
  window_data = all_time_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "reversed",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

all_time_reversed_summary <- summarise_counterfactual(
  simulations = all_time_reversed_counterfactual$simulations,
  observed_alignment = all_time_observed_alignment,
  calibration = all_time_reversed_counterfactual$calibration
)
```

```{r all-time-calibration-table}
bind_rows(
  all_time_active_counterfactual$calibration,
  all_time_reversed_counterfactual$calibration
) %>%
  select(
    year,
    calibration_metric,
    n_candidates,
    n_kept,
    n_nonzero_attention,
    n_zero_attention,
    n_unique_nonzero_condition_sets,
    observed_calibration_ci,
    mean_kept_calibration_ci,
    median_abs_error,
    max_abs_error,
    calibration_quality,
    seed
  ) %>%
  arrange(calibration_metric, year)
```

```{r all-time-active-plot, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(
  all_time_active_summary,
  metric = "active",
  x_label = "Burden year",
  title = "All-time attention across burden years: active CI"
)
```

```{r all-time-active-count-table}
summarise_null_compatibility(all_time_active_summary, metric = "active") %>%
  arrange(location_name)
```

```{r all-time-active-table}
all_time_active_summary %>%
  filter(
    location_name != "Global",
    year %in% c(min(years), 2010, 2015, 2020, max(years))
  ) %>%
  select(
    location_name,
    year,
    observed_ci,
    sim_lci,
    sim_median,
    sim_uci,
    observed_minus_sim_median,
    observed_percentile_rank,
    null_compatibility
  ) %>%
  arrange(location_name, year)
```

```{r all-time-reversed-plot, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(
  all_time_reversed_summary,
  metric = "reversed",
  x_label = "Burden year",
  title = "All-time attention across burden years: reversed CI"
)
```

```{r all-time-reversed-count-table}
summarise_null_compatibility(all_time_reversed_summary, metric = "reversed") %>%
  arrange(location_name)
```

```{r all-time-reversed-table}
all_time_reversed_summary %>%
  filter(
    location_name != "Global",
    year %in% c(min(years), 2010, 2015, 2020, max(years))
  ) %>%
  select(
    location_name,
    year,
    observed_reversed_ci,
    sim_reversed_lci,
    sim_reversed_median,
    sim_reversed_uci,
    observed_reversed_minus_sim_median,
    observed_reversed_percentile_rank,
    reversed_null_compatibility
  ) %>%
  arrange(location_name, year)
```

## Interpretation of Results

The four analyses separate two related questions:

```text
active CI:
  Is GWAS attention concentrated on high-burden conditions?

reversed CI:
  Is disease burden concentrated among conditions that receive GWAS attention?

time-varying attention:
  What do specific GWAS attention windows do?

all-time attention:
  What does the accumulated GWAS canon do?
```

For time-varying attention, lower-SDI regions tend to fall below the
High-SDI-calibrated null for both active and reversed concentration-index
definitions. This indicates that individual GWAS attention windows do not merely
reflect passive transfer from High SDI burden priorities. They perform worse for
lower-SDI burden than a generic High-SDI-calibrated attention allocation would.
Substantively, this suggests that the observed changing GWAS canon is a
particular slice of the High-SDI-compatible allocation space: it is concentrated
on conditions that are less relevant to lower-SDI burden than alternative
allocations that would still reproduce observed High SDI alignment.

For all-time attention, the reversed metric is closer to the null expectation.
This suggests that the accumulated GWAS literature has some coverage of
conditions carrying lower-SDI disease burden. In other words, lower-SDI burden
is not entirely outside the studied condition universe once GWAS attention is
pooled across time.

However, the all-time active metric again suggests weaker lower-SDI alignment
than expected under the High-SDI transfer null. This means that even where
lower-SDI-relevant conditions have received some GWAS attention, the amount of
attention is not concentrated on the highest-burden lower-SDI conditions as much
as the null would predict.

Taken together, the results suggest a distinction between coverage and
prioritisation intensity. GWAS has accumulated some condition coverage relevant
to lower-SDI disease burden, but the distribution of attention remains more
skewed toward a narrow subset of High-SDI-compatible conditions than expected
under passive High-SDI transfer. The evidence does not support concerted
lower-SDI-responsive condition prioritisation, which would appear as lower-SDI
alignment above the null ribbon. Instead, the stronger pattern is that observed
lower-SDI alignment is often worse than expected even under a
High-SDI-calibrated null. The compatibility-count tables above quantify this
pattern by reporting how many years fall below, inside, or above the null ribbon
for each SDI region.

This interpretation should be weighted by calibration quality. Windows flagged
with calibration warnings are useful descriptively, but stronger claims should
focus on well-calibrated windows or be revisited with a larger candidate count
and tuned proposal sampler.
