---
title: "High-SDI alignment counterfactual"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
source(here::here("scripts/regional_analysis/R/packages.R"))
source(here::here("scripts/regional_analysis/R/data_loaders.R"))
source(here::here("scripts/regional_analysis/R/metrics.R"))
source(here::here("scripts/regional_analysis/R/plots.R"))
source(here::here("scripts/simulation-sdi-response/high_sdi_counterfactual.R"))

simulation_seed <- 20260615
n_candidates <- 5000
n_keep <- 500
median_abs_error_warn <- 0.02
max_abs_error_warn <- 0.05

# Paper-scale starting point:
# n_candidates <- 50000
# n_keep <- 1000
```

## Aim

This analysis tests whether lower-SDI attention-burden alignment is compatible
with a High-SDI transfer null. The counterfactual preserves each sliding
window's observed attention sparsity and inequality, then keeps simulated
attention allocations closest to the observed High SDI attention-burden
concentration index.

The candidate generator uses a proposal mixture. One component assigns observed
nonzero attention scores fully at random; the other components assign the same
observed scores to randomly selected conditions with noisy High-SDI rank
alignment. The proposal mixture is used to explore enough High-SDI-aligned
allocations for calibration; kept allocations are selected by closeness to the
observed High SDI concentration index.

The active and reversed metrics are calibrated in separate simulation runs. The
active ribbon keeps allocations closest to observed High SDI attention-burden
concentration index; the reversed ribbon keeps allocations closest to observed
High SDI reversed concentration index.

COVID-19 is excluded before building observed and simulated windows so a single
pandemic-era condition does not dominate the range of possible permutations.

The all-time sensitivity uses the all-time attention score distribution and
evaluates it against each burden year.

## Data

```{r prepare-data}
window_data <- prepare_counterfactual_windows(exclude_covid = TRUE)
observed_alignment <- compute_observed_counterfactual_alignment(window_data)

years <- sort(unique(window_data$year))

tibble(
  first_year = min(years),
  last_year = max(years),
  n_years = length(years),
  n_conditions = n_distinct(window_data$cause_id),
  n_locations = n_distinct(window_data$location_name)
)
```

```{r observed-summary}
observed_alignment %>%
  filter(
    sex_name == "Both",
    year %in% c(min(years), 2010, 2015, 2020, max(years)),
    location_name %in% c("High SDI", "Low SDI", "Global")
  ) %>%
  select(
    location_name,
    year,
    concentration_index,
    concentration_index_rev,
    attention_gini,
    alignment_ratio,
    share_alignment,
    n_causes
  ) %>%
  arrange(location_name, year)
```

## Simulation

```{r run-counterfactual}
active_counterfactual <- run_high_sdi_counterfactual(
  window_data = window_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "active",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

active_counterfactual_summary <- summarise_counterfactual(
  simulations = active_counterfactual$simulations,
  observed_alignment = observed_alignment,
  calibration = active_counterfactual$calibration
)

reversed_counterfactual <- run_high_sdi_counterfactual(
  window_data = window_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "reversed",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

reversed_counterfactual_summary <- summarise_counterfactual(
  simulations = reversed_counterfactual$simulations,
  observed_alignment = observed_alignment,
  calibration = reversed_counterfactual$calibration
)
```

```{r calibration-summary}
bind_rows(
  active_counterfactual$calibration,
  reversed_counterfactual$calibration
) %>%
  select(
    year,
    calibration_metric,
    n_candidates,
    n_kept,
    n_nonzero_attention,
    n_zero_attention,
    n_unique_nonzero_condition_sets,
    observed_high_sdi_ci,
    observed_high_sdi_reversed_ci,
    mean_kept_high_sdi_ci,
    mean_kept_high_sdi_reversed_ci,
    observed_calibration_ci,
    mean_kept_calibration_ci,
    median_abs_error,
    max_abs_error,
    calibration_quality,
    seed
  ) %>%
  arrange(calibration_metric, year)
```

Windows are flagged when `median_abs_error` exceeds
`r median_abs_error_warn` or `max_abs_error` exceeds
`r max_abs_error_warn`. A warning marks a window as poorly calibrated for strong
interpretation; it does not remove the window from the plot.

## Observed vs null ribbon

The red line and points show the observed metric. The blue line shows the
median of the High-SDI-calibrated null distribution; the blue ribbon shows its
2.5th to 97.5th percentile range.

```{r null-ribbon-active, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(active_counterfactual_summary, metric = "active")
```

```{r null-summary-table-active}
active_counterfactual_summary %>%
  filter(
    year %in% c(min(years), 2010, 2015, 2020, max(years)),
    location_name != "Global"
  ) %>%
  select(
    location_name,
    year,
    observed_ci,
    sim_lci,
    sim_median,
    sim_uci,
    observed_minus_sim_median,
    observed_percentile_rank,
    null_compatibility
  ) %>%
  arrange(location_name, year)
```

## Reversed observed vs null ribbon

The reversed metric asks whether disease burden is concentrated among
conditions with greater GWAS attention. It is included as a diagnostic rather
than the primary attention-burden alignment metric.

```{r null-ribbon-reversed, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(reversed_counterfactual_summary, metric = "reversed")
```

```{r null-summary-table-reversed}
reversed_counterfactual_summary %>%
  filter(
    year %in% c(min(years), 2010, 2015, 2020, max(years)),
    location_name != "Global"
  ) %>%
  select(
    location_name,
    year,
    observed_reversed_ci,
    sim_reversed_lci,
    sim_reversed_median,
    sim_reversed_uci,
    observed_reversed_minus_sim_median,
    observed_reversed_percentile_rank,
    reversed_null_compatibility
  ) %>%
  arrange(location_name, year)
```

## Calibration check

High SDI is the calibration target. Its observed line should sit close to the
kept permutation distribution by construction; lower-SDI regions are the actual
test of the transfer null.

```{r high-sdi-active-calibration-plot, fig.width=7, fig.height=4}
plot_counterfactual_ribbon(
  active_counterfactual_summary %>% filter(location_name == "High SDI"),
  include_global = TRUE,
  metric = "active"
)
```

```{r high-sdi-reversed-calibration-plot, fig.width=7, fig.height=4}
plot_counterfactual_ribbon(
  reversed_counterfactual_summary %>% filter(location_name == "High SDI"),
  include_global = TRUE,
  metric = "reversed"
)
```

## Effective sparsity

```{r sparsity-plot, fig.width=7, fig.height=4}
active_counterfactual$calibration %>%
  ggplot(aes(x = year, y = n_nonzero_attention)) +
  geom_line() +
  geom_point() +
  theme_report() +
  labs(
    x = "GWAS attention window",
    y = "Nonzero attention conditions"
  )
```

## All-time attention sensitivity

This sensitivity repeats the counterfactual using all-time GWAS attention scores
and each year of DALYs. It asks whether the accumulated GWAS condition canon is
narrower than a generic High-SDI-calibrated canon across the burden history.
Because the all-time attention distribution is dense and highly unequal,
calibration warnings should be checked before placing strong weight on these
ribbons under the default candidate count.

```{r all-time-counterfactual}
all_time_data <- prepare_counterfactual_all_time(
  burden_years = years,
  exclude_covid = TRUE
)

all_time_observed_alignment <- compute_observed_counterfactual_alignment(all_time_data)

all_time_active_counterfactual <- run_high_sdi_counterfactual(
  window_data = all_time_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "active",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

all_time_active_summary <- summarise_counterfactual(
  simulations = all_time_active_counterfactual$simulations,
  observed_alignment = all_time_observed_alignment,
  calibration = all_time_active_counterfactual$calibration
)

all_time_reversed_counterfactual <- run_high_sdi_counterfactual(
  window_data = all_time_data,
  years = years,
  n_candidates = n_candidates,
  n_keep = n_keep,
  seed = simulation_seed,
  calibration_metric = "reversed",
  median_abs_error_warn = median_abs_error_warn,
  max_abs_error_warn = max_abs_error_warn
)

all_time_reversed_summary <- summarise_counterfactual(
  simulations = all_time_reversed_counterfactual$simulations,
  observed_alignment = all_time_observed_alignment,
  calibration = all_time_reversed_counterfactual$calibration
)
```

```{r all-time-calibration-table}
bind_rows(
  all_time_active_counterfactual$calibration,
  all_time_reversed_counterfactual$calibration
) %>%
  select(
    year,
    calibration_metric,
    n_candidates,
    n_kept,
    n_nonzero_attention,
    n_zero_attention,
    n_unique_nonzero_condition_sets,
    observed_calibration_ci,
    mean_kept_calibration_ci,
    median_abs_error,
    max_abs_error,
    calibration_quality,
    seed
  ) %>%
  arrange(calibration_metric, year)
```

```{r all-time-active-plot, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(all_time_active_summary, metric = "active")
```

```{r all-time-active-table}
all_time_active_summary %>%
  filter(
    location_name != "Global",
    year %in% c(min(years), 2010, 2015, 2020, max(years))
  ) %>%
  select(
    location_name,
    year,
    observed_ci,
    sim_lci,
    sim_median,
    sim_uci,
    observed_minus_sim_median,
    observed_percentile_rank,
    null_compatibility
  ) %>%
  arrange(location_name, year)
```

```{r all-time-reversed-plot, fig.width=11, fig.height=4.5}
plot_counterfactual_ribbon(all_time_reversed_summary, metric = "reversed")
```

```{r all-time-reversed-table}
all_time_reversed_summary %>%
  filter(
    location_name != "Global",
    year %in% c(min(years), 2010, 2015, 2020, max(years))
  ) %>%
  select(
    location_name,
    year,
    observed_reversed_ci,
    sim_reversed_lci,
    sim_reversed_median,
    sim_reversed_uci,
    observed_reversed_minus_sim_median,
    observed_reversed_percentile_rank,
    reversed_null_compatibility
  ) %>%
  arrange(location_name, year)
```
