---
title: "GWAS Attention Score Comparison"
author: ""
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

## Introduction

This document compares the old and new GWAS attention scores to assess changes in the methodology and their impact on disease prioritization.

## Load Libraries

```{r libraries}
library(here)
library(dplyr)
library(data.table)
library(rineq)
library(ggplot2)
```

## Load Data

We load both the new and old attention score datasets.

```{r load-data}
# Load the new dataset
gwas_attention_new <- fread(here("Data/merged_dataset_exclude_Injuries_2023.csv")) %>%
  select(cause_name = `Cause Name`, cause_id, total_attention_score) %>%
  filter(!duplicated(cause_id))

# Load the old dataset for comparison
gwas_attention_old <- fread(here("Data/Ncase_merged_dataset_exclude_Injuries.csv")) %>%
  select(cause_name = `Cause Name`, cause_id, total_attention_score) %>%
  filter(!duplicated(cause_id))
```

## Merge Datasets

```{r merge-data}
merged <- merge(gwas_attention_new, gwas_attention_old, by = "cause_id", suffixes = c("_new", "_old"))
merged
```

## Correlation Analysis

Calculate the Pearson correlation between old and new attention scores:

```{r correlation}
cor(merged$total_attention_score_new, merged$total_attention_score_old, method = "pearson")
```

This is really high - but it's because there are some really large values -

```{r scatter-full}
ggplot(merged, aes(x = total_attention_score_old, y = total_attention_score_new)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Correlation between Old and New Attention Scores",
       x = "Old Attention Score",
       y = "New Attention Score") +
  theme_minimal()
```


Examining the relationship for smaller attention scores:

```{r scatter-zoomed}
ggplot(merged, aes(x = total_attention_score_old, y = total_attention_score_new)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Correlation between Old and New Attention Scores (Zoomed In)",
       x = "Old Attention Score",
       y = "New Attention Score") +
  theme_minimal() +
  xlim(0, 1e7) +
  ylim(0, 1e7)
```

There are a number of new attention scores that are zero but were previously non-zero - Identify diseases those diseases

```{r new-diseases-count}
sum(merged$total_attention_score_new == 0 & merged$total_attention_score_old > 0)

```

List of studies that now have zero attention scores

```{r new-diseases-list}
merged %>%
  filter(total_attention_score_new == 0 & total_attention_score_old > 0) %>%
  select(cause_name_new, cause_id)
```
