---
title: "Attention Score Pipeline Comparison"
subtitle: "_updated_6 reference vs regenerated pipeline output"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(here)
source(here("scripts/attention_scores/config.R"))

cfg <- load_project_config()
before_path  <- cfg_path(cfg_get(cfg, "inputs.reference_attention_scores"), cfg)
after_path   <- cfg_path(cfg_get(cfg, "attention_scores.temporal_output"), cfg)
manual_path  <- cfg_path(cfg_get(cfg, "inputs.manual_zero_mapping"), cfg)
todo_path    <- cfg_path(cfg_get(cfg, "pipeline_checks.zero_attention_todo"), cfg)

before <- fread(before_path) %>% mutate(source = "updated_6_reference")
after  <- fread(after_path)  %>% mutate(source = "pipeline_current")

combined <- bind_rows(before, after)
key_cols <- c("cause_name", "cause_id", "analysis_type", "time_strata")

# Terms already covered in the manual mapping file (have at least one filled URI)
manual_mapped <- fread(manual_path, na.strings = c("", "NA")) %>%
  filter(!is.na(mapped_trait_uri), nzchar(mapped_trait_uri)) %>%
  distinct(cause_name)

# Current pipeline zeros not yet manually mapped → write todo file
a_all_for_todo <- after %>% filter(analysis_type == "all")
todo <- a_all_for_todo %>%
  filter(total_attention_score == 0,
         !cause_name %in% manual_mapped$cause_name) %>%
  left_join(
    before %>% filter(analysis_type == "all") %>%
      select(cause_name, reference_score = total_attention_score),
    by = "cause_name"
  ) %>%
  arrange(desc(coalesce(reference_score, 0L))) %>%
  transmute(cause_name, cause_id, reference_score,
            mapped_trait_uri = NA_character_, notes = NA_character_)
dir.create(dirname(todo_path), recursive = TRUE, showWarnings = FALSE)
fwrite(todo, todo_path, na = "")
```

## Summary

This report compares Rayan's manually generated `_updated_6` file with the
regenerated pipeline output. The current pipeline reads the configured
`canonical_mapping.master_mapping` path
(generated by `scripts/canonical_mapping/export_to_master_mapping.R` from the
LLM/human-reviewed evidence package) and rolls up through the GBD hierarchy.

```{r summary-table}
b_all <- before %>% filter(analysis_type == "all")
a_all <- after  %>% filter(analysis_type == "all")

gained <- inner_join(
  b_all %>% select(cause_name, cause_id, before_score = total_attention_score),
  a_all %>% select(cause_name, after_score  = total_attention_score),
  by = "cause_name"
)

tibble(
  metric = c(
    "All-time causes (total)",
    "All-time zero-score (_updated_6)",
    "All-time zero-score (pipeline)",
    "Reference-zero causes recovered by pipeline",
    "Reference nonzero causes with higher pipeline score"
  ),
  value = c(
    nrow(a_all),
    sum(b_all$total_attention_score == 0),
    sum(a_all$total_attention_score == 0),
    sum(gained$before_score == 0 & gained$after_score > 0),
    sum(gained$before_score > 0  & gained$after_score > gained$before_score)
  )
)
```

## Reference-Zero Causes Recovered by Pipeline

These GBD terms had zero all-time attention in `_updated_6` and now have non-zero
scores in the regenerated pipeline output.

```{r zero-to-nonzero}
gained %>%
  filter(before_score == 0, after_score > 0) %>%
  arrange(desc(after_score)) %>%
  select(cause_name, cause_id, before_score, after_score)
```

## Score Changes for Shared Non-Zero Causes

```{r nonzero-changes}
changed <- gained %>%
  filter(before_score > 0, after_score != before_score) %>%
  mutate(
    delta = after_score - before_score,
    pct_change = round(100 * delta / before_score, 1)
  ) %>%
  arrange(desc(abs(delta)))

cat("Causes with score change:", nrow(changed), "\n")
changed %>%
  select(cause_name, cause_id, before_score, after_score, delta, pct_change) %>%
  head(20)
```

## File Shape

```{r file-shape}
combined %>%
  group_by(source) %>%
  summarise(
    rows = n(),
    unique_causes = n_distinct(cause_name),
    zero_all_time = sum(analysis_type == "all" & total_attention_score == 0),
    nonzero_all_time = sum(analysis_type == "all" & total_attention_score > 0),
    score_sum_all_time = sum(total_attention_score[analysis_type == "all"], na.rm = TRUE),
    .groups = "drop"
  )
```

## Temporal Row Counts

```{r temporal-shape}
combined %>%
  filter(analysis_type != "all") %>%
  group_by(source, analysis_type) %>%
  summarise(
    rows = n(),
    nonzero = sum(total_attention_score > 0),
    .groups = "drop"
  ) %>%
  arrange(analysis_type, source)
```

## Score Distribution (All-Time, Non-Zero)

```{r score-distribution}
bind_rows(
  b_all %>% filter(total_attention_score > 0) %>%
    select(source, cause_name, score = total_attention_score),
  a_all %>% filter(total_attention_score > 0) %>%
    select(source, cause_name, score = total_attention_score)
) %>%
  ggplot(aes(x = score, fill = source)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  scale_x_log10() +
  scale_fill_manual(values = c(updated_6_reference = "#E69F00", pipeline_current = "#56B4E9")) +
  labs(x = "All-time attention score (log10)", y = "Count", fill = "Pipeline") +
  theme_minimal()
```

## Before vs After Score Scatter (All-Time)

```{r scatter}
gained %>%
  filter(before_score > 0 | after_score > 0) %>%
  ggplot(aes(x = before_score + 0.1, y = after_score + 0.1, label = cause_name)) +
  geom_abline(slope = 1, intercept = 0, color = "grey60", linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Reference score (_updated_6; log10, +0.1)",
    y = "Pipeline score (log10, +0.1)",
    title = "All-time score: _updated_6 vs regenerated pipeline"
  ) +
  theme_minimal()
```

## Zero in Pipeline but Non-Zero in Reference (`_updated_6`)

These causes have zero attention in the current pipeline but had non-zero scores in
`_updated_6`. They are **missing mappings** — the canonical mapping has no accepted
term for them that matches any GWAS Catalog publication.

```{r zero-pipeline-nonzero-ref}
problematic <- gained %>%
  filter(after_score == 0, before_score > 0) %>%
  arrange(desc(before_score))

cat("Causes zero in pipeline but non-zero in _updated_6:", nrow(problematic), "\n")
problematic %>%
  select(cause_name, cause_id, after_score, before_score)
```

```{r zero-pipeline-nonzero-ref-plot}
if (nrow(problematic) > 0) {
  problematic %>%
    mutate(cause_name = reorder(cause_name, before_score)) %>%
    ggplot(aes(x = before_score, y = cause_name)) +
    geom_col(fill = "#D55E00") +
    scale_x_continuous(labels = scales::comma) +
    labs(
      x = "Reference score (_updated_6)",
      y = NULL,
      title = "Causes with zero pipeline score but non-zero reference score"
    ) +
    theme_minimal(base_size = 9)
}
```

## Remaining Unmapped Zero-Attention Traits

Terms with zero pipeline attention that are **not yet covered** in
`manual_zero_attention_mapping.csv`. This list is also written to the configured
`pipeline_checks.zero_attention_todo` path — add new mappings there, then copy the
`cause_name` + `mapped_trait_uri` rows into `manual_zero_attention_mapping.csv`
and re-run the pipeline.

```{r todo}
cat("Remaining unmapped zero-attention traits:", nrow(todo), "\n")
todo %>% select(cause_name, cause_id, reference_score)
```

## All Zero-Attention Traits in Pipeline

All GBD universe terms with zero all-time attention in the current pipeline,
regardless of their reference score.

```{r all-zero-pipeline}
all_zero <- a_all %>%
  filter(total_attention_score == 0) %>%
  left_join(
    b_all %>% select(cause_name, reference_score = total_attention_score),
    by = "cause_name"
  ) %>%
  arrange(desc(coalesce(reference_score, 0L))) %>%
  select(cause_name, cause_id, reference_score)

cat("Total zero-attention traits in pipeline:", nrow(all_zero), "\n")
all_zero
```

## Score Agreement on Shared All-Time Keys

```{r score-agreement}
inner_join(
  before %>% filter(analysis_type == "all") %>%
    select(all_of(key_cols), before_score = total_attention_score),
  after  %>% filter(analysis_type == "all") %>%
    select(all_of(key_cols), after_score  = total_attention_score),
  by = key_cols
) %>%
  summarise(
    n = n(),
    correlation = cor(before_score, after_score, use = "complete.obs"),
    causes_unchanged = sum(before_score == after_score),
    causes_increased = sum(after_score > before_score),
    causes_decreased = sum(after_score < before_score)
  )
```

## Key Overlap

```{r key-overlap}
before_keys <- before %>% select(all_of(key_cols))
after_keys  <- after  %>% select(all_of(key_cols))

data.frame(
  metric = c("keys_in_both", "keys_only_before", "keys_only_after"),
  rows = c(
    nrow(inner_join(before_keys, after_keys, by = key_cols)),
    nrow(anti_join(before_keys, after_keys, by = key_cols)),
    nrow(anti_join(after_keys, before_keys, by = key_cols))
  )
)
```
